A family of accessibility measures derived from spatial interaction principles

### 2025-04-06 paper/manuscript.qmd
comment on section two divergent research streams

The spatial-interaction-focused stream of literature inspired by @wilson1971 that cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable. Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p. 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976]. Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model). Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating]. Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].

On the other hand, we discern that the potential-for-spatial-interaction-focused stream of literature inspired by @hansen1959 that cite both @hansen1959 and @wilson1971, cite @wilson1971 for three general reasons: - Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining]. - Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]. While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction. This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]. - Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. However, as outlined in preceding sections, we argue the resulting indicator values are plagued by similar interpretability issues. Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways for maintaining constraints in opportunities in the context of competitive accessibility.

>>>>

The stream of literature focused on spatial interaction models inspired by @wilson1971 and which cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable. Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p. 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976]. Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model). Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating]. Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].

On the other hand, we discern that the stream of literature focused on accessibility inspired by @hansen1959 cites @wilson1971 for three general reasons: - Firstly, as attribution for using using different context-dependent travel cost functions <!-- no need to cite all of these references here. Keep it to 4 --> [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining]. - Secondly, associating spatial interaction with accessibility's potential for spatial interaction <!-- no need to cite all of these references here. Keep it to 4 --> [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]. While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction. This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]. - Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts <!-- no need to cite all of these references here. Keep it to 4 --> [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. However, as outlined in preceding sections, we argue the resulting indicator values are plagued by similar interpretability issues. Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways for maintaining constraints in opportunities in the context of competitive accessibility.

<!-- check text of the paragraphp above for clarity -->

<!-- if only soukhov et al 2023 and 2024 have previously used 'balancing factors as ways for maintaining constraints', it might, you might wanna explain in more detail what was dones in those papers. this is also important to position the contribution of this current paper compared to those previous studies -->
